A Survey of Quantitative Descriptions of Molecular Structure

### 2012-03-16 paper.tex
Added a chunk of text and started added references

\section{Running Notes}
\begin{itemize}
\item What are descriptors, why are they needed?
\item Survey of descriptor types, focusing on nature of inputs,
  complexity, interpretabilty 
\item Talk about biological descriptors (see paper from Tropsha that
  uses bioactivity data as descriptors)
\item Discuss how fingerprints can also be used as descriptors
\item Also mention descriptors for non-small molecule cases
  \begin{itemize}
  \item proteins
  \item polymers
  \item inorganics (maybe ask Maciej for pointers)
  \end{itemize}
\item Survey of descriptor implementations (OSS and non-OSS)
  \begin{itemize}
  \item MOE
  \item DRAGON
  \item CDK
  \item RDKit
  \item JChem
  \item PipelinePilot
  \item OEChem/GraphTK
  \item JOELib
  \item CODESSA
  \item Schrodinger stuff (Meastro?)
  \end{itemize}
\item Go into more detail of OSS descriptor implementations
  \begin{itemize}
  \item CDK (describe architecture), mention wrappers around the
    library (CDKDescUI and batch mode)
  \item PaDEL
  \end{itemize}
\item Compare OSS descs to commercial ones - point to the Pfizer/CDD
  paper that compared MOE and CDK
\end{itemize}
\clearpage
\newpage

@@

>>>>

\section{A Categorization of Descriptors}
\label{sec:categ-descr}

We have noted that a molecular structure can be characterized using a
numebr of numerical descriptors. In this section we describe a
categorization of descriptors. Admittedly, the grouping is somewhat
arbitrary but serves as a useful guide. In addition, the
categorization also considers the nature of the chemical structure
being considered - some descriptors are only useful when applied to
small molecules, whereas other descriptors are defined specifically
for polymers or protein structures.

A final class of descriptors are fingerprints. Traditionally, these
descriptors are represented in the form of bit strings. Binary
fingerprints can be divided into hashed fingerprints, where
substructures (such as paths of length $n$) are converted to a string
representation and then hashed into a randomly selected bit positions,
or keyed fingerprints, where each bit position corresponds to a unique
substructural feature. The nature of the features can range from
simple functional groups (hydroxyl, carbonyl) and topological
substructures (paths, chains, cycles) to atom environments
\cite{Bremser1978,Bender:2004aa} and pharmacophoric elements
\cite{Renner:2006aa}.

While the descriptors mentioned so far are derived from some form of
chemical structure, an alternative approach is to employ observed
biological activities of a molecule as descriptors of that molecule.
This approach was taken by Sedykh et al \cite{Sedykh:2011fk}   who
hypothesized that dose response data points from high throughput
dose-reponse assays could be employed as biological descriptors. They
observed that when such descriptors were combined with traditional
chemical structure descriptors, the predictive performance of models
developed using the combined set was better than those developed using
conventional approaches. A similar approach can be taken using the
PASS methodology \cite{Poroikov:2007aa}, though this is an indirect
method - one must predict the PASS profile from the chemical
structure, which can then be used in subsequent predictive models [REF?].

\section{Descriptor Implementations}
\label{sec:descr-impl}

A variety of descriptor implementations are available, with most
cheminformatics toolkits providing implementations of a common set of
descriptors and fingerprints. Table \ref{tab:impl} summarizes the
descriptor classes implemented by several toolkits and applications.

Given the multiple implementations of many descriptors, it is natural
to compare their performance. As noted previously, it is desirable,
but unlikely that implementations of the same descriptor using
different cheminformatcs toolkits will be identical. Gupta et al
employed SMARTS based descriptors from the CDK and MOE to develop
decision tree models to predict human liver microsomal metabolic
stability\cite{Gupta:2010uq}. Their results indicated very similar
performance between the two implementations.

One can also directly compare descriptor values generated using
different implementations.

A number of physicochemical descriptors commonly employed in
predictive modeling are themselves models of an experimental
property. For example, log P can be experimentally measured and a
number of algorithms have been developed to predict log P
[REFs]. Given the utility of log P in drug discovery, a number of
implementations are available. Given that this descriptor is a
surrogate for an actual experimental property, it is reasonable to
compare calculated log P values from different implementations to the
experimentall observed values, rather than between themselves. Figure
\ref{fig:logp} compares computed log P values from the CDK, ChemAxon
and ACD Labs for a set of 10,000 molecules taken from the logPstar
dataset [REF].

\section{Running Notes}
\begin{itemize}
\item \textbf{Done} What are descriptors, why are they needed?
\item Survey of descriptor types, focusing on nature of inputs,
  complexity, interpretabilty 
\item \textbf{Done} Talk about biological descriptors (see paper from Tropsha that
  uses bioactivity data as descriptors)
\item Discuss how fingerprints can also be used as descriptors
\item Also mention descriptors for non-small molecule cases
  \begin{itemize}
  \item proteins
  \item polymers
  \item inorganics (maybe ask Maciej for pointers)
  \end{itemize}
\item Survey of descriptor implementations (OSS and non-OSS)
  \begin{itemize}
  \item MOE
  \item DRAGON
  \item CDK
  \item RDKit
  \item JChem
  \item PipelinePilot
  \item OEChem/GraphTK
  \item JOELib
  \item CODESSA
  \item Schrodinger stuff (Meastro?)
  \end{itemize}
\item Go into more detail of OSS descriptor implementations
  \begin{itemize}
  \item CDK (describe architecture), mention wrappers around the
    library (CDKDescUI and batch mode)
  \item PaDEL
  \end{itemize}
\item Compare OSS descs to commercial ones - point to the Pfizer/CDD
  paper that compared MOE and CDK
\end{itemize}
\clearpage
\newpage

@@

\ctable[caption={A comparison of Topological Polar Surface Area values
generated using the CDK and ACD Labs software, for 57,600 molecules
taken from Pubchem AID 1996},cap={},label={fig:tpsa},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{tpsa-cdk-acd}
}

\ctable[caption={A comparison of experimental versus calculated logP values
generated using the CDK, ACD Labs and ChemAxon software, for 10,000 molecules
taken from the proprietary logPstar dataset [REF?]},cap={},label={fig:logp},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{logp-comp-exptl}
}
